---
title: "Shorebird Survey"
author: Alex S.
output: 
  html_document:
    number_sections: no
    theme: journal
    toc: yes
    toc_float:
      collapsed: no
  pdf_document:
    toc: yes
date: "Last Updated: `r format(Sys.time(), '%Y-%m-%d')`" 
---

```{r setup, include=FALSE, echo=FALSE, cache=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      include = FALSE,
                      cache = TRUE,
                      warning = FALSE, 
                      message = FALSE)

library(tidyverse)
library(RColorBrewer)
library(lubridate)
library(readxl)
library(broom)

# Utils
g <- function(.) glimpse(.)

medium_fonts <- ggplot2::theme_bw() +
  ggplot2::theme(
    plot.title = ggplot2::element_text(size = 14),
    axis.text.x = ggplot2::element_text(size = 12),
    axis.text.y = ggplot2::element_text(size = 12),
    axis.title.x = ggplot2::element_text(size = 12),
    axis.title.y = ggplot2::element_text(size = 12),
    # legend.title = ggplot2::element_blank(),
    legend.text = ggplot2::element_text(size = 12),
    strip.text.x = ggplot2::element_text(size = 12),
    strip.text.y = ggplot2::element_text(size = 12)
  )
```


# Overview

Basic EDA of shorebird data.

Goals:
1) Is there enough data to make statements about changes in species over time? i.e. can we get statistical significance for differences between Aramburu & Pickelweed.
2) Has the number of species or abundance of any species changed over time on Aramburu vs Pickleweed?
3) Are there differences between high/low tide?
4) Are there obvious improvements to methodology we can make?

Qs:
- is Pickleweed our control? Y
- were both high and low tide measured for the same amount of time?
- for a particular observation day it looks like each bird can be in there for every isle, tide, position, behabior, and observer(?)
- generally ever sampled day is sampled at every tide and at all locations??


# Load & format data

From Excel file:

ISLE CODES:   
AS = ARAMBURU, SOUTH END.   
AC = ARAMBURU CENTRAL.   
AN = ARAMBURU, NORTH END.    
U = UN-NAMED.   
P = PICKLEWEED.

POSITION CODES:  
V=VEGETATION.  
RS=ROCK/SAND.  
M=MUDFLAT.  
SW=STANDING IN WATER.   
FW=FLOATING ON WATER.  
OW=OVER WATER

BEHAVIOR CODES:  
R=ROOSTING/RESTING.  
F=FORAGING (probing, dabbling, diving, aerial surveying).  
O =OTHER (Comment)


```{r}
island_key <- data.frame(
  short_name = c("as", "ac", "an", "u", "p"),
  long_name = c("aramburu_south", "aramburu_central", "aramburu_north", "unnamed", "pickleweed")
)

position_key <- data.frame(
  short_name = c("v", "rs", "m", "sw", "fw", "ow"),
  long_name = c("vegetation", "rock_sand", "mudflat", "standing_in_water", "floating_on_water", "over_water")
)

behavior_key <- data.frame(
  short_name = c("r", "f", "o"),
  long_name = c("roosting", "foraging", "other")
)

dat <- read_excel("data/SHOREBIRD_DATA_Complete_2018.xls", sheet = 1) %>% 
  rename_all(tolower) %>%
  select(-analyze, -`pre/post construction`) %>%
  rename(obs_date = date,
         tide_ft = `tide (ft)`,
         tide_hl = `high/low`, # I assume this is tide?
         island = `aram?`,
         aou_num = `aou num`,
         record_num = `record num`,
         gulls = gullls
         ) %>% 
  
  mutate(obs_date = as.Date(obs_date),
         month_no = lubridate::month(obs_date),
         year = lubridate::year(obs_date) # there were many missing year entries, populate here
         ) %>% 

  # species names
  mutate(species = tolower(species)) %>% 
  mutate(species = ifelse(species %in% c("none", "-", "0", NA), "none", species)) %>% 
  mutate(is_bird = !(species %in% c("deer", "humans"))) %>% 
  
  # convert guild to logical
  mutate(shorebirds = grepl("x", tolower(shorebirds)),
         peeps = grepl("x", tolower(peeps)),
         gulls = grepl("x", tolower(gulls)),
         terns = grepl("x", tolower(terns)),
         wading = grepl("x", tolower(wading)),
         grebes = grepl("x", tolower(grebes))
         ) %>% 
  
  # tide 
  mutate(tide_hl = gsub('"', "", tolower(tide_hl))) %>% 
  # mutate(tide_hl = factor(tide_hl, levels = "low", "mid", "high")) %>%
  
  # remove some funny data (TODO: check with Julia)
  subset(isle != "ROAMING" & isle != "*") %>% 
  
  # fill missing location data
  mutate(island = ifelse(tolower(isle) %in% c("as", "ac", "an"), "Aramburu", 
                         ifelse(tolower(isle) %in% c("p"), "Pickleweed", "Unknown")))
  
dat %>% g


```

# Basic EDA

## Sampling Methods

Observations
- typically all chunks of the island were surveyed on every observation date
- typically measured at both high and low tide, sometimes only one. Occasionaly also "mid" tide.
- data spans 2009-2018, with 1-7 observations per month
- no obs in 1H 2009, 2H 2012, 2H 2015, and all of 2017

### per island
Looks like survey included all island areas virtually every time

```{r}
dat %>% 
  group_by(obs_date, isle) %>% 
  summarize(n_sp = n()) %>% 
  mutate(has_count = !is.na(n_sp)) %>% 
  ggplot(aes(x = factor(obs_date), y = isle, fill = has_count)) +
  geom_raster() +
  scale_x_discrete(breaks = NULL) +
  medium_fonts +
  labs(x = "observation date", y = "island")
```

### tide
Looks like survey included all island areas virtually every time

```{r}
dat %>% 
  group_by(obs_date, tide_hl) %>% 
  summarize(n_sp = n()) %>% 
  mutate(has_count = !is.na(n_sp)) %>% 
  ggplot(aes(x = factor(obs_date), y = tide_hl, fill = has_count)) +
  geom_raster() +
  scale_x_discrete(breaks = NULL) +
  medium_fonts +
  labs(x = "observation date", y = "tide")
```


### per month
```{r}
dat %>% 
  select(isle, year, month_no, obs_date) %>% 
  distinct() %>% 
  group_by(isle, year, month_no) %>% 
  summarize(n_obs = n()) %>% 
  ungroup() %>%
  # complete missing observations:
  tidyr::spread(month_no, n_obs) %>% 
  tidyr::gather(month_no, n_obs, 3:15) %>% 
  mutate(has_count = !is.na(n_obs)) %>% 
  mutate(month_no = as.numeric(month_no)) %>% 
  subset(!is.na(month_no)) %>%
  ggplot(aes(x = month_no, y = year, fill = n_obs)) +
  geom_raster(aes(fill = n_obs)) +
  facet_wrap(~isle) +
  scale_x_continuous(breaks = seq(1,12)) +
  scale_y_continuous(breaks = seq(2009,2018)) +
  theme_bw() +
  labs(title = "Observations per Month")

```

## Species counts
```{r}
dat %>% 
  subset(isle != "*") %>% 
  subset(tide_hl != "mid") %>% 
  group_by(obs_date, isle, tide_hl) %>% 
  summarize(n_species = n(),
            n_individuals = sum(qty)) %>% 
  ggplot(aes(x = obs_date, y = n_species, group = tide_hl)) +
  geom_point(aes(color = tide_hl), size = 1, alpha = 0.8) +
  geom_smooth(aes(color = tide_hl)) +
  medium_fonts +
  facet_wrap(~isle)

dat %>% 
  subset(island != "Unknown") %>% 
  subset(tide_hl != "mid") %>% 
  group_by(obs_date, island, tide_hl) %>% 
  summarize(n_species = n(),
            n_individuals = sum(qty)) %>% 
  ggplot(aes(x = obs_date, y = n_species, group = tide_hl)) +
  geom_point(aes(color = tide_hl), size = 1, alpha = 0.8) +
  geom_smooth(aes(color = tide_hl)) +
  medium_fonts +
  facet_wrap(~island)
```

## Individual counts
```{r}
dat %>% 
  subset(isle != "*") %>% 
  subset(tide_hl != "mid") %>% 
  group_by(obs_date, isle, tide_hl) %>% 
  summarize(n_species = n(),
            n_individuals = sum(qty)) %>% 
  ggplot(aes(x = obs_date, y = n_individuals, group = tide_hl)) +
  geom_point(aes(color = tide_hl), size = 1, alpha = 0.8) +
  geom_smooth(aes(color = tide_hl)) +
  medium_fonts +
  facet_wrap(~isle) +
  ylim(0,800)

dat %>% 
  subset(island != "Unknown") %>% 
  subset(tide_hl != "mid") %>% 
  group_by(obs_date, island, tide_hl) %>% 
  summarize(n_species = n(),
            n_individuals = sum(qty)) %>% 
  ggplot(aes(x = obs_date, y = n_individuals, group = tide_hl)) +
  geom_point(aes(color = tide_hl), size = 1, alpha = 0.8) +
  geom_smooth(aes(color = tide_hl)) +
  medium_fonts +
  facet_wrap(~island) +
  ylim(0,1000)
```

## Species low  vs high tide

Difference in species observed, high vs. low tide.

```{r}
# number of individuals per species, per date, isle, tide (grouped across observers, behaviors, positions, comments, etc.)
tide_qty_per_sp <- dat %>% 
  subset(tide_hl %in% c("low", "high")) %>%
  subset(is_bird) %>% 
  group_by(obs_date, island, tide_hl, common) %>% 
  summarize(qty = sum(qty, na.rm = T))

tide_qty_per_sp %>% 
  group_by(obs_date, island, tide_hl) %>%
  summarize(n_species_observed = n()) %>%
  tidyr::spread(tide_hl, n_species_observed) %>% 
  mutate(difference = low-high) %>% 
  ggplot(aes(difference)) +
  geom_vline(xintercept = 0, linetype = 2) +
  geom_bar(alpha = 0.9) +
  facet_wrap(~island) +
  medium_fonts +
  labs(x = "difference", y = "number of sampling days", 
       title = "High vs. Low Tide difference in # species observed", 
       subtitle = "positive values: low tide has more species", 
       caption = "(paired comparison)")
```


Estimate whether difference in frequency of observation of each species is significantly different between high and low tide.

```{r}
tide_qty <- tide_qty_per_sp %>% 
  group_by(island, tide_hl, common) %>% 
  summarize(n_times_sp_observed = n()) %>% 
  # fill missing values with 0s:
  tidyr::spread(tide_hl, n_times_sp_observed) %>% # number times species was observed
  replace_na(list(high = 0, low = 0)) %>% 
  tidyr::gather(tide_hl, n_times_sp_observed, high, low) %>% 
  left_join(
    tide_qty_per_sp %>% 
      group_by(island, tide_hl) %>% 
      summarize(n_surveyed = length(unique(obs_date))),
    by = c("island", "tide_hl")
  ) 

tide_diffs <- tide_qty %>% 
  group_by(island, common) %>%
  arrange(tide_hl) %>% 
  do(tidy(prop.test(.$n_times_sp_observed, .$n_surveyed))) %>% 
  rename(high = estimate1,
         low = estimate2)  %>% 
  mutate(difference = low - high) # difference in whether bird was sighted on a survey between low & high tide
```

```{r}
plot_dat <- tide_diffs %>% 
  # subset(island != "Unknown") %>%
  subset(island == "Aramburu") %>%
  subset(!is.na(common)) %>% 
  subset(p.value < 0.01) %>% 
  select(island, common, low, high) %>% 
  mutate(diff_type = ifelse(low > high, "more at low tide", "fewer at low tide")) %>% 
  tidyr::gather(tide_hl, frequency, low, high) %>% 
  mutate(tide_hl = factor(tide_hl, levels = c("low", "high"))) %>% 
  mutate(tide_hl_num = as.numeric(tide_hl)) 

plot_dat %>% 
  arrange(diff_type)

plot_dat %>% 
  ggplot(aes(x = tide_hl_num, y = frequency)) +
  geom_line(aes(color = common)) +
  geom_point(aes(color = common), size = 3) +
  geom_text(data = subset(plot_dat, tide_hl_num == 1), aes(label = common, color = common), 
            size = 3, hjust = 1.15, fontface = "bold") +
   geom_text(data = subset(plot_dat, tide_hl_num == 2), aes(label = common, color = common), 
             size = 3, hjust = -0.15, fontface = "bold") +
  facet_wrap(~diff_type) + 
  medium_fonts +
  theme(strip.text.x = ggplot2::element_text(size = 10), 
        legend.position = "none") +
  scale_x_continuous(limits = c(0.0, 3), breaks = c(1,2), labels = c("low", "high"), minor_breaks = NULL) +
  labs(x = "Tide", y = "Frequency Species is Observed (per survey)", 
       title = "Species with significantly different presence at low vs. high tide",
       caption = "p < 0.01, chi-squared test")

```


# Change over time

## Species Count